---
title: "In class assignment - week 9"
author: "Suncica Milosevic"
date: "10/20/2020"
output: 
  html_document:
      toc: TRUE
      toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Here is a demonstration on how to use R Markdown:

## Making a new chunk

The shortcut key for windows is Ctrl-Alt-i

On mac Cmnd-Option-i

I press this command, markdown will insert a new R code chunk.

If I want, I can give each chunk a name and it will appear in the navigation pane. 

```{r example chunk}

```


## Reading a csv file

I'm going to read some data from a csv (comma-separated-values) file.

I can use read.csv() along with the path to the file to read the data into a data.frame.

Here's an example:



```{r read_birds_data}
require(here)
dat_birds = read.csv(here("data", "hab.sta.csv"))
#comments
```

## Previewing the contents of a data.frame

I can use the head() function to print the first few lines of a data.frame.

Here's an example with the birds data

```{r head function demo}
head(dat_birds)


```

## R demo plot

I can use histogram:

When I plot this data, the histogram loos weird.

```{r R demo plot}
hist(dat_birds$aspect )
```
##How to customize a histogram

I can title by using main ="   "

```{r custom histogram}
hist(dat_birds$aspect, 
     main = "Histogram of study site aspect",
     breaks = 36)
     
```



